Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-05 Paper/grass.tex
pull from shreyas
Added the emphasis on threshold condition in intro

In this paper we present a mathematical \shreyas model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear instability of a uniform flow over the grass bed. 
The threshold condition for the onset of the instability and the resulting waving frequency compares well with experimental observations.
In the limit of a dense grass bed, two distinct unstable modes can be asymptotically distinguished, with one of them sharing many characteristics with the Kelvin-Helmholtz instability. 
In the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

@@

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.

@@

@@

In conclusion we note that linear stability analysis of flow in presence of canopy reveals two distinct modes of flow instability which might lead to monami. 
The occurrence of these modes for flow instability in a canopy depends on its submergence ratio and shear layer thickness. Our analysis predict a threshold criteria characterized by a Reynolds number. 
The predicted value of threshold Reynolds number agrees well with experimentally observed parameters. 
Inclusion of variation of drag coefficient along the grass filament might lead to better agreement between the observed and predicted frequency of monami. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.
\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left[({\rho U_0 H}/\mu) (C_N d N_g H)\right]^{-1/3}$} \\
 Linearized dynamics	& Inviscid Rayleigh's equation			& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey$, $\delta/H$ 	& $\Rey \delta^{3/2}$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}

>>>>

In this paper we present a mathematical model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is a primary consideration that sets the threshold flow.
We deduce this numerically and through an asymptotic analysis of the dense grass limit.

@@

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.
The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$; departures from this scaling can be seen in the inset of figure \ref{basicflow}.

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left[({\rho U_0 H}/\mu) (C_N d N_g H)\right]^{-1/3}$} \\
 Linearized dynamics	& Inviscid Rayleigh's equation			& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey$, $\delta/H$ 	& $\Rey \delta^{3/2}$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}

@@

In conclusion we note that linear stability analysis of flow in presence of canopy reveals two distinct modes of flow instability which might lead to monami. 
The occurrence of these modes for flow instability in a canopy depends on its submergence ratio and shear layer thickness. Our analysis predict a threshold criteria characterized by a Reynolds number. 
The predicted value of threshold Reynolds number agrees well with experimentally observed parameters. 
Inclusion of variation of drag coefficient along the grass filament might lead to better agreement between the observed and predicted frequency of monami. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.
